Automated detection of data fabrication using statistical tools

### 2017-07-27 submission/manuscript.Rmd
Revisions in progress

---
title: "Ecological performance of detecting data fabrication"
author: "Chris HJ Hartgerink, Jan G Voelkel, Jelte M Wicherts, Marcel ALM van Assen"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_depth: 2
csl: ../bibliography/apa.csl
bibliography: ../bibliography/library2.bib
---

```{r preparation, echo = FALSE}
suppressPackageStartupMessages(if(!require(pROC)){install.packages('pROC', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(pROC))
suppressPackageStartupMessages(if(!require(knitr)){install.packages('knitr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(if(!require(foreign)){install.packages('foreign', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(foreign))
suppressPackageStartupMessages(if(!require(latex2exp)){install.packages('latex2exp', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(latex2exp))
suppressPackageStartupMessages(if(!require(ggplot2)){install.packages('ggplot2', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(if(!require(gridExtra)){install.packages('gridExtra', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(if(!require(xtable)){install.packages('xtable', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(xtable))
suppressPackageStartupMessages(if(!require(httr)){install.packages('httr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(httr))
suppressPackageStartupMessages(if (!require(car)){install.packages('car', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(if (!require(effects)){install.packages('effects', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(effects))
suppressPackageStartupMessages(if (!require(lsr)){install.packages('lsr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(lsr))
suppressPackageStartupMessages(if (!require(data.table)){install.packages('data.table', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(if (!require(dplyr)){install.packages('dplyr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(if (!require(reshape2)){install.packages('reshape2', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(if (!require(metafor)){install.packages('metafor', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(metafor))

@@

Any field of empirical inquiry is faced with cases of scientific misconduct at some point, either in the form of fabrication, falsification or plagiarism (FFP). Psychology was faced with Stapel; medical sciences were faced with Poldermans and Macchiarini; life sciences were faced with Voignet. These are just a few examples of misconduct cases in the last decade. Overall, an estimated 2\% of all scholars have admitted to falsifying or fabricating research results at least once [@10.1371/journal.pone.0005738] and this is likely to be an underestimate due to socially desirable responses. The detection rate of data fabrication is likely to be even lower; for example, only around a dozen cases are discovered in the United States and the Netherlands, despite covering several hundreds of thousands of researchers. At best, this amounts to a detection rate far below 1\% of those 2\% who admit to fabricating data --- the tip of a seemingly much larger iceberg.

In order to stifle attempts at data fabrication, improved detection of fabricated data is considered to deter such harmful attempts. 
Although deterrence theory dates back to the middle of the 17th century [@leviathan], its implementation has not occurred equally across fabrication, falsification, and plagiarism. Basically, deterrence theory stipulates that with increased risk of detection, the utility of scientific misconduct (for this context) will decrease and therefore fewer people will engage in such behaviors. This principle of deterrence has been implemented with plagiarism scanners, a development that already started a long time ago [e.g., @10.1109/13.28038]. In the last decade, detecting image manipulation has become one of the few forms of detecting data fabrication . The Journal of Cell Biology and EMBO scan each submitted image for potential manipulation [@The_Journal_of_Cell_Biology2015-vh;@10.1038/546575a], which greatly increases the risk of detecting (blatant) image manipulation. More recently, algorithms have been developed to automate the scanning of images for (subtle) manipulations [@10.1007/s11948-016-9841-7]. These developments in detecting image manipulation have increased detection risk during the pre-publication and post-publication phase by improving detection mechanisms and increasing the understanding of how images might be manipulated. Moreover, their application also helps researchers systematically evaluate research articles to estimate the extent of the problem of image manipulation [4\% of all papers are estimated to contain manipulated images, @10.1128/mBio.00809-16].

Statistical methods provide one way to improve detection of data fabrication in empirical research. Humans are notoriously bad at understanding and estimating probabilities [e.g., @10.1126/science.185.4157.1124;@10.1037/h0031322], which could manifest itself in the fundamentally probabilistic data they try to fabricate. 
When data are fabricated, probabilistic principles are easily violated if forgotten at the univariate level, bivariate level, trivatiate level, or beyond [@Haldane1948-nm]. Based on this idea, statistical methods that investigate whether the reported data are feasible under the theoretically probabilistic processes can be used to detect potential data fabrication.

```{r echo=FALSE}
set.seed(1234)
ex1 <- t(replicate(n = 10, expr = table(sample(x = c("M", "F"), size = 100, replace = TRUE))))
ex1p <- pchisq(apply(ex1, 1, function(x) sum((x-50)^2/50)), df = 1, lower.tail = FALSE)
set.seed(1234)
ex2_1 <- 50 + sample(x = -1:1, size = 10, replace = TRUE)
ex2_2 <- 100 - ex2_1
ex2 <- cbind(ex2_1, ex2_2)
ex2p <- pchisq(apply(ex2, 1, function(x) sum((x-50)^2/50)), df = 1, lower.tail = FALSE)

df <- data.frame(study = sprintf("Study %s", 1:10),
                 ex1,
                 ex1p,
                 ex2,
                 ex2p)
names(df) <- c("Study",
               "Experimental [S1]",
               "Control [S1]",
               "P-value [S1]",
               "Experimental [S2]",
               "Control [S2]",
               "P-value [S2]")
```

The application of such statistical methods to detect data fabrication has occurred in several cases of scientific misconduct in recent years and has potential for future application beyond these specific cases. 
For example, problems in papers by Fuji were highlighted with statistical methods [@10.1111/j.1365-2044.2012.07128.x;@10.1111/anae.13126], resulting in 183 retractions [@oransky2015]. This method works as follows: in true randomized clinical trials (RCTs), baseline measurements should be statistically equal across groups. As such, the $p$-values for group comparisons would be expected to be uniformly distributed because the null hypothesis of equal groups is true by definition of the design. In the Fuji papers, group comparisons showed excessive consistency, resulting primarily in high $p$-values (e.g., .99, .95) and a high mean $p$-value across the comparisons, where a mean $p$-value of .5 is expected. As an illustration, see Table `r tabnr`, which depicts the assignment of participants to the experimental or control condition for a true randomized design (Set 1) or for a fabricated design (Set 2). The mean $p$-value for the true randomized design Set 1 is `r round(mean(ex1p), 3)`, whereas the fabricated Set 2 has mean $p$-value `r round(mean(ex2p), 3)`. Other statistical methods to detect data fabrication are addressed in the theoretical framework.
.

```{r table_excess, echo = FALSE}
knitr::kable(df, digits = 3, caption = "Examples of participant counts for genuine- and fabricated randomized clinical trials. Set 1 (S1) is randomly generated data under the null hypothesis of random assignment (assumed to be the genuine process), whereas Set 2 (S2) is generated under excessive consistency with equal groups. Each trial contains 100 participants. The $p$-values ")

tabnr <- tabnr + 1
```

Such statistical methods to detect data anomalies, although developed to quantify suspicions in a specific paper, could be applied to screen multiple papers. The application of such methods can be (semi-)automated if data are available in a machine-readable format that one of the statistical methods can be applied to. An example of such a potential case for mass application of using statistics to detect (potential) data fabrication is in the ClinicalTrials.gov database, where baseline measures across randomized groups are readily available for download and subsequent analysis [@10.3897/rio.1.e7462].

Nonetheless, prior to applying statistical methods to flag potentially problematic results, investigating whether such methods function well enough in detecting problems is required for responsible application. 
We hardly know how researchers might go about fabricating data. Cases such as Stapel, Fuji, Smeesters, and Sanna provide some insights, but are highly pre-selected (i.e., those who got caught/confessed) and as such, systematically biased. Relatively extensive descriptions in rare and partial autobiogrophical accounts provide little insight into the actual data fabrication process, except for the setting where it might take place [e.g., late at night when no one is around; @stapel_book]. However, trustworthiness of these accounts can be called into question. Additionally, the performance of methods to detect data fabrication is highly dependent on the unknown prevalence of data fabrication and the power to actually to detect data fabrication. Given that we do not know how researchers might fabricate data, the diagnosticity of these methods cannot realistically be assessed.

Throughout this paper, we inspect statistical methods to detect potential data fabrication that can be applied to (1) summary results or (2) raw data. 
Even though structure and contents of data can look different depending on the structure of a study and the measures, there are certain common characteristics of empirical results and the underlying raw data that can be inspected. For example, summary results frequently include means, standard deviations, test-statistics, and $p$-values. Raw data frequently contain at least some variables measured at a interval- or ratio scale [@10.1126/science.103.2684.677]. Such common characteristics allow for the development of generic statistical methods that can be applied across a varied set of results to screen for problematic data. We review the theoretical framework of the specific methods we apply throughout this paper, but these are not exhaustive of all methods available to test for potential problems in empirical data [see also @buyse1999;@10.7287/peerj.preprints.2400v1;@10.7287/peerj.preprints.2064v1;@sprite-heathers].

@@

In the current paper, we differentiate between statistical methods to detect potential data fabrication based on reported data (i.e., summary results) or raw data. For summary results, we review $p$-value analysis, variance analysis, and effect size analysis as potential ways to detect data anomalies. $P$-value analysis can be applied whenever sufficient nonsignificant $p$-values are reported; variance analysis can be applied whenever a set of variances are reported for independent groups alongside the sample sizes per group; effect size analysis can be used whenever the effect size can be computed [e.g., an APA reported $t$- or $F$-statistic;@10.1525/collabra.71].  For raw data, we review digit analyses (i.e., the Newcomb-Benford law and terminal digit analysis), and multivariate associations as potential ways to detect data anomalies. Digit analyses can be applied when ratio scale measures are present in the raw data; multivariate associations can be applied whenever there is a multivariate relation in the raw data that can be computed and control data is available from the literature.

## Detecting data fabrication in summary results

@@

The distribution of a $p$-value is uniform if the null hypothesis is true and right-skewed if the alternative hypothesis is true [@fisher1925]. One $p$-value is the result of the population effect size, the precision of the estimate, and the observed effect size, whose properties carry over to a set of $p$-values if those $p$-values are independent. As such, the $p$-value distribution of a set of $p$-values is uniform when the null hypothesis is true, or right-skewed when the alternative hypothesis is true.

When the $p$-value distribution is not uniform nor right-skewed, it can indicate potential data anomalies. For example, a distribution might show a bump when excessive $p$-hacking occurs that seeks barely significant results [e.g., Figure 1 in @10.7717/peerj.1935]. However, data fabricators might create other non-uniform or non-right-skewed distributions, failing to take into account the theoretical boundary conditions. For example, given the frequent misinterpretation of $p$-values [@10.3758/BF03213921;@10.1007/s11336-015-9446-0;@10.7326/0003-4819-130-12-199906150-00008], it is possible that the same underlying misinterpration will manifest itself in fabricated data. More specifically, given the frequent misinterpretation of $p$-valuesas the probability of an effect [@10.1053/j.seminhematol.2008.04.003;@10.1136/bmj.311.7003.485], researchers who fabricate nonsignificant data might forget to fabricate a uniform $p$-value distribution and generate only large $p$-values when fabricating nonsignificant results.

In order to test whether observed $p$-values are anomalous, we proposed an adaptation of Fisher's method [@10.1186/s41073-016-0012-9]. This adaptation is a simple reversal of the original Fisher method [@fisher1925], which was introduced as a meta-analytic test for the presence of an effect. This test is computed as 
$$
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(p_i)
$$
where it tests for right-skew (i.e., more smaller $p$-values than larger $p$-values) across the $k$ number of $p$-values. Reversing the Fisher method results in 
$$
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(1-\frac{p_i-t}{1-t})
$$
where it now tests for left-skew (i.e., more larger $p$-values than smaller $p$-values) across the $k$ number of $p$-values that falls above the threshold $t$. This threshold is added and can be set to any value between zero and one. Upon writing this paper, it became clear to us that this is reversed Fisher method is similar to the operating principle of Carlisle's method testing for excessive homogeneity across baseline measurements in RCTs [@10.1111/anae.13938;@10.1111/j.1365-2044.2012.07128.x;@10.1111/anae.13126].

Despite this test being useful for detecting data anomalies in nonsignificant $p$-values, one exception should be taken into account: wrongly specified one-tailed tests. For properly specified one-tailed tests, the $p$-value distribution is right-skewed. When wrongly specified, this distribution is reversed and becomes left-skew. As such, any data anomalies detected with this method would need to be inspected for misspecified one-tailed hypotheses to preclude false conclusions.

@@

Variance- or standard deviation estimates are typically reported to indicate dispersion, but just like the mean there should be sampling error in this estimate proportional to the sample size [i.e., $\sigma/\sqrt{2n}$ under the assumption of normality, p. 351;@yule1922]. A variance estimate follows a $\chi^2$-distribution, which is dependent on the sample size [p. 445; @hogg-tanis]; that is
$$
z^2_j\sim\left(\frac{\chi^2_{N_j-1}}{N_j-1}\right)/MS_w
$$
where $N_j$ is the sample size of the $j$th group and $MS_w$ is the normalizing constant resulting in a standardized variance $z^2_j$. The normalizing constant $MS_w$ is computed as 
$$
MS_w=\frac{\sum\limits^k_{j=1}(N_j-1)s^2_j}{\sum\limits^k_{j=1}(N_j-1)}
$$
where $s^2_j$ is the variance in the $j$th group.

The observed dispersion of the variances can be compared to various measures of expected dispersion in the variances. Dispersion can be operationalized in various ways, such as the standard deviation of the variances [denoted in this paper as $SD_z$;@10.1177/0956797613480366] or as the range of the variances (denoted as $max-min_z$). Too consistent results would indicate potential anomalies in the reported data. For example, in the Smeesters case three independent conditions from the same study ($n_k=15$) were reported to have standard deviations 25.09, 24.58, and 25.65. The standard deviation of the standard deviations here is `r round(sd(c(25.09, 24.58, 25.65)), 2)` (i.e., $SD_z$). Such consistency (or more consistency) would only be observed in `r round(std_var(n = rep(15, 3), sds = c(25.09, 24.58, 25.65), iter = 100000)*100, 2)`% of 100,000 simulated replications [@10.1177/0956797613480366].

>>>>

---
title: "Automated detection of data fabrication using statistical tools"
author: "Chris HJ Hartgerink, Jan G Voelkel, Jelte M Wicherts, Marcel ALM van Assen"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
 pdf_document:
  toc: yes
 html_document:
  toc: yes
  toc_depth: 2
csl: ../bibliography/apa.csl
bibliography: ../bibliography/library2.bib
---

# Abstract

PLACEHOLDER

## Keywords

PLACEHOLDER

```{r preparation, echo = FALSE}
suppressPackageStartupMessages(if(!require(pROC)){install.packages('pROC', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(pROC))
suppressPackageStartupMessages(if(!require(knitr)){install.packages('knitr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(if(!require(foreign)){install.packages('foreign', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(foreign))
suppressPackageStartupMessages(if(!require(latex2exp)){install.packages('latex2exp', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(latex2exp))
suppressPackageStartupMessages(if(!require(ggplot2)){install.packages('ggplot2', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(if(!require(gridExtra)){install.packages('gridExtra', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(if(!require(xtable)){install.packages('xtable', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(xtable))
suppressPackageStartupMessages(if(!require(httr)){install.packages('httr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(httr))
suppressPackageStartupMessages(if (!require(car)){install.packages('car', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(if (!require(lsr)){install.packages('lsr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(lsr))
suppressPackageStartupMessages(if (!require(data.table)){install.packages('data.table', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(if (!require(dplyr)){install.packages('dplyr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(if (!require(reshape2)){install.packages('reshape2', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(if (!require(metafor)){install.packages('metafor', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(metafor))

@@

# Introduction

Any field of empirical inquiry is faced with cases of scientific misconduct at some point, either in the form of fabrication, falsification or plagiarism (FFP). Psychology was faced with Stapel; medical sciences were faced with Poldermans and Macchiarini; life sciences were faced with Voignet. These are just a few examples of misconduct cases in the last decade. Overall, an estimated 2\% of all scholars have admitted to falsifying or fabricating research results at least once [@10.1371/journal.pone.0005738], which is likely to be an underestimate due to socially desirable responses. The detection rate of data fabrication is likely to be even lower; for example, only around a dozen cases become public in the United States and the Netherlands, despite covering several hundreds of thousands of researchers. At best, this suggests a detection rate far below 1\% of those 2\% who admit to fabricating data --- the tip of a seemingly much larger iceberg.

In order to stifle attempts at data fabrication, improved detection of fabricated data is considered to deter such harmful attempts. 
This idea is based on deterrence theory [@leviathan], which stipulates that with increased risk of detection, the utility of scientific misconduct will decrease and therefore fewer people will engage in such behaviors. Implementation of deterrence has not occurred equally across fabrication, falsification, and plagiarism. For plagiarism, it has been implemented with plagiarism scanners, a development that already started a long time ago [e.g., @10.1109/13.28038]. For data fabrication, detecting image manipulation has increasingly become possible and implemented. The Journal of Cell Biology and the EMBO journal scan each submitted image for potential manipulation [@The_Journal_of_Cell_Biology2015-vh;@10.1038/546575a], which greatly increases the risk of detecting (blatant) image manipulation. More recently, algorithms have been developed to automate the scanning of images for (subtle) manipulations [@10.1007/s11948-016-9841-7]. These developments in detecting image manipulation have increased detection risk during the pre-publication and post-publication phase and increase the understanding of how images might be manipulated. Moreover, their application also helps researchers systematically evaluate research articles to estimate the extent of the problem of image manipulation [4\% of all papers are estimated to contain manipulated images, @10.1128/mBio.00809-16].

Statistical methods provide one way to improve detection of data fabrication in empirical research. Humans are notoriously bad at understanding and estimating randomness [@10.1126/science.185.4157.1124;@10.1037/h0031322;@10.1037/1082-989X.5.2.241], which could manifest itself in the fundamentally probabilistic data they try to fabricate. 
When data are fabricated, principles of statistics and randomness are easily violated at the univariate level, bivariate level, trivatiate level, or beyond [@Haldane1948-nm]. Based on this idea, statistical methods that investigate whether the reported data are feasible under the theoretically probabilistic processes can be used to detect potential data fabrication.

```{r echo=FALSE}
m <- 50
sd <- 10
n <- 100

set.seed(1234)
x1_1 <- replicate(n = 10, expr = rnorm(n, m, sd))
x1_2 <- replicate(n = 10, expr = rnorm(n, m, sd))

ex1_1em <- apply(x1_1, 2, mean)
ex1_1esd <- apply(x1_1, 2, sd)
ex1_1cm <- apply(x1_2, 2, mean)
ex1_1csd <- apply(x1_2, 2, sd)
ex1_1tp <- NULL
for (i in 1:dim(x1_1)[2]) ex1_1tp[i] <- t.test(x = x1_1[,i], y = x1_2[,i], var.equal=TRUE)$p.value
ex1_1tp <- unlist(ex1_1tp)

set.seed(1234)
x2_1em <- replicate(n = 10, expr = m + runif(1, 0, 20))
x2_2cm <- replicate(n = 10, expr = m + runif(1, 0, 20))
x2_1esd <- replicate(n = 10, expr = sd + runif(1, 0, 1.5))
x2_2csd <- replicate(n = 10, expr = sd + runif(1, 0, 1.5))

sdpooled <- ((n - 1) * x2_1esd^2 + (n - 1) * x2_2csd^2)/(n * 2 - 2)
tval <- (x2_1em - x2_2cm) / sdpooled
ex2_1tp <- pt(abs(tval), n * 2 - 2, lower.tail = FALSE) * 2

# Just saving for later use in theory section
# set.seed(1234)
# std_var(n = rep(100, length(x2_1esd) * 2), sds = c(x2_1esd, x2_2csd), iter = 10000)

df <- data.frame(study = sprintf("Study %s", 1:10),
         sprintf("%s (%s)", round(ex1_1em, 3), round(ex1_1esd, 3)),
         sprintf("%s (%s)", round(ex1_1cm, 3), round(ex1_1csd, 3)),
         ex1_1tp,
         sprintf("%s (%s)", round(x2_1em, 3), round(x2_1esd, 3)),
         sprintf("%s (%s)", round(x2_2cm, 3), round(x2_2csd, 3)),
         ex2_1tp)
names(df) <- c("Study",
        "$M_E$ ($SD_E$) [S1]",
        "$M_C$ ($SD_C$) [S1]",
        "P-value [S1]",
        "$M_E$ ($SD_E$) [S2]",
        "$M_C$ ($SD_C$) [S2]",
        "P-value [S2]")
```

Statistical methods to detect data fabrication have been applied in several cases of scientific misconduct in recent years and has potential for future application beyond these specific cases. 
For example, excessive consistency in papers by Fuji were highlighted with statistical methods [@10.1111/j.1365-2044.2012.07128.x;@10.1111/anae.13126], resulting in 183 retractions [@oransky2015]. Carlisle's method works as follows: in true randomized clinical trials (RCTs) baseline measurements should be identically distributed across groups. As such, the $p$-values for group comparisons would be expected to be uniformly distributed because the null hypothesis of identical distributions across groups is true by definition of the randomized design. In the Fuji papers, group comparisons showed excessive consistency, resulting primarily in high $p$-values (e.g., .99, .95) and a high mean $p$-value across the comparisons, where a mean $p$-value of .5 is expected. As an illustration, see Table `r tabnr`, which depicts 10 hypothetical studies containing 100 participants per condition, for true randomized designs (Set 1) or for fabricated designs (Set 2). The mean $p$-value for the true randomized design Set 1 is `r round(mean(ex1_1tp), 3)`, whereas the fabricated Set 2 has mean $p$-value `r round(mean(ex2_1tp), 3)`. Other statistical methods to detect data fabrication are addressed in the theoretical sections.
.

```{r table_excess, echo = FALSE}
knitr::kable(df, digits = 3, caption = "Examples of means and standard deviations for a continuous outcome in genuine- and fabricated randomized clinical trials. Set 1 (S1) is randomly generated data under the null hypothesis of random assignment (assumed to be the genuine process), whereas Set 2 (S2) is generated under excessive consistency with equal groups. Each trial condition contains 100 participants. The $p$-values are the result of independent $t$-tests comparing the experimental and control conditions within each respective set.")
```

Statistical methods to detect data fabrication, although developed to quantify suspicions in a specific paper, could be applied to screen multiple papers. The application of such methods can be (semi-)automated if data are available in a machine-readable format that one of the statistical methods can be applied to. An example of such a potential case for mass application of using statistics to detect (potential) data fabrication is in the ClinicalTrials.gov database, where baseline measures across randomized groups are readily available for download and subsequent analysis [@10.3897/rio.1.e7462].

Nonetheless, prior to applying statistical methods to flag potentially fabricated results, investigating whether such methods perform well enough in detecting data fabrication is required for responsible application. 
We hardly know how researchers might go about fabricating data. Cases such as Stapel, Fuji, Smeesters, and Sanna provide some insights, but are highly pre-selected (i.e., those who got caught/confessed) and as such, may be systematically biased. Relatively extensive descriptions in rare and partial autobiogrophical accounts provide little insight into the actual data fabrication process, except for the setting where it might take place [e.g., late at night when no one is around; @stapel_book]. Additionally, the performance of methods to detect data fabrication is highly dependent on the unknown prevalence of data fabrication and the power to actually detect data fabrication. Given that we do not know how researchers might fabricate data, the diagnosticity of these methods cannot realistically be assessed.

Throughout this paper, we inspect statistical methods to detect potential data fabrication that can be applied to summary statistics (Study 1) or raw data (Study 2). 
Even though structure and contents of data can look different depending on the structure of a study and the measures, there are certain common characteristics of empirical results and the underlying raw data that can be inspected. For example, summary statistics frequently include means, standard deviations, test-statistics, and $p$-values. Raw data frequently contain at least some variables measured at a interval- or ratio scale [@10.1126/science.103.2684.677]. Such common characteristics allow for the development of generic statistical methods that can be applied across a varied set of results to screen for fabricated data. For each study, we review the theory of the specific methods we apply. However, the reviewed methods are not exhaustive of all methods available to test for potential data fabrication in empirical data [see also, @buyse1999;@10.7287/peerj.preprints.2400v1;@10.7287/peerj.preprints.2064v1;@sprite-heathers].

@@

In the current paper, we differentiate between statistical methods to detect potential data fabrication based on reported summary statistics or raw data. For summary statistics, we review $p$-value analysis, variance analysis, and effect size analysis as potential ways to detect data fabrication. $P$-value analysis can be applied whenever sufficient nonsignificant $p$-values are reported; variance analysis can be applied whenever a set of variances are reported for independent groups alongside the sample sizes per group; effect size analysis can be used whenever the effect size is reported or can be computed [e.g., an APA reported t- or F-statistic;@10.1525/collabra.71]. For raw data, we review digit analyses (i.e., the Newcomb-Benford law and terminal digit analysis) and multivariate associations as potential ways to detect data fabrication. Both types of digit analyses can be applied when ratio scale measures are present in the raw data and only terminal digit analysis can be applied when there are continuous measures. Multivariate associations can be investigated whenever there are two or more variables and data on that same relation is available from (assumably) genuine data sources.

## Detecting data fabrication in summary statistics

@@

The distribution of a $p$-value is uniform if the null hypothesis is true and right-skewed if the alternative hypothesis is true [@fisher1925]. The distribution of one $p$-value is the result of the population effect size, the precision of the estimate, and the observed effect size, whose properties carry over to a set of independent $p$-values if those $p$-values are independent. As such, the $p$-value distribution of a set of $p$-values is uniform when the null hypothesis is true, or right-skewed when the alternative hypothesis is true.

When a $p$-value distribution of independent $p$-values is not uniform or right-skewed, as would be theoretically expected, it can indicate potential data fabrication. For example, in the Fuji case, supposedly identically distributed groups were fabricated. However, this resulted in excessively large $p$-values and ultimately the identification of potential data fabrication by @10.1111/j.1365-2044.2012.07128.x. Our illustration in Table `r tabnr` also shows the difference between expected data under the null distribution (Set 1) and what might result from excessively consistent and fabricated data (Set 2). More specifically, the expected value of a uniform $p$-value distribution is .5, but the fabricated data from our illustration result in a mean $p$-value of `r round(mean(ex2_1tp), 3)`.

In order to test whether the distribution of nonsignificant and independent $p$-values might be fabricated, we proposed an adaptation of Fisher's method [@10.1186/s41073-016-0012-9]. This adaptation is a reversal of the original Fisher method [@fisher1925], which was introduced as a meta-analytic test for the presence of an effect. This test is computed as 
$$F
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(p_i)
$$
where it tests for more smaller $p$-values than larger $p$-values across the $k$ number of $p$-values. Reversing the original Fisher method, which tests for right-skew, into a Fisher method that tests for left-skew results in 
$$
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(1-\frac{p_i-t}{1-t})
$$
where it now tests for more larger $p$-values than smaller $p$-values across the $k$ number of $p$-values that fall above the threshold $t$. When the threshold is set to zero, it selects all $p$-values, but when this $t>0$ the remaining $p$-values are rescaled to fit the original 0-1 range (i.e., by dividing by $1-t$). Upon writing this paper, it became clear to us that this reversed Fisher method is similar to the operating principle of Carlisle's method testing for excessive homogeneity across baseline measurements in RCTs [@10.1111/anae.13938;@10.1111/j.1365-2044.2012.07128.x;@10.1111/anae.13126], which we had not realized before.

```{r, echo = FALSE}
threshold <- .05
genuine_p <- ex1_1tp[ex1_1tp > threshold]
fab_p <- ex2_1tp[ex2_1tp > threshold]
genuine_chi <- -2 * sum(log(1 - ((genuine_p - threshold) / (1 - threshold))))
fab_chi <- -2 * sum(log(1 - ((fab_p - threshold) / (1 - threshold))))
genuine_chi_p <- pchisq(q = genuine_chi, df = 2 * length(genuine_p), lower.tail = FALSE)
fab_chi_p <- pchisq(q = fab_chi, df = 2 * length(fab_p), lower.tail = FALSE)
x <- c(.21, -.08, -.37, -.08, .32)
pval <- pnorm(abs(x), 0, 1, lower.tail = FALSE) * 2
pval <- pval[pval > threshold]
chi <- -2 * sum(log(1 - ((pval - threshold) / (1 - threshold))))
p_chi <- pchisq(q = chi, df = 2 * length(pval), lower.tail = FALSE)
```

As an example, we apply the reversed Fisher method to both the genuine- and fabricated results from our illustration in Table `r tabnr` and data from the Fuji case. Using the threshold $t=`r threshold`$ to only select the nonsignificant results, we retain $k=`r length(genuine_p)`$ genuine $p$-values and $k=`r length(fab_p)`$ fabricated $p$-values. Calculating the $\chi^2$ value for sets of $k$ nonsigificant $p$-values results in a $\chi^2=`r round(genuine_chi, 3)`$ for the $p$-values resulting from the genuine data and $\chi^2=`r round(fab_chi, 3)`$ for the $p$-values resulting from the fabricated data, with $p$-values `r round(genuine_chi_p, 3)` and `r round(fab_chi_p, 3)` respectively, using $2k$ as the degrees of freedom of the $\chi^2$-test. Another example, directly from the Fuji case [@10.1111/j.1365-2044.2012.07128.x], anecdotally illustrates that fabricated data can result in significant findings with the reversed Fisher method. $p$-values from five independent comparisons for one dependent measure [fentanyl dose, as extracted from Table 3 in @10.1111/j.1365-2044.2012.07128.x] show excessively high $p$-values, $\chi^2_{`r length(pval) * 2`}=`r round(chi, 3)`, p=`r round(p_chi, 3)`$.

Despite this test being useful for detecting data fabrication in nonsignificant $p$-values, one exception should be taken into account: wrongly specified one-tailed tests. For properly specified one-tailed tests, the $p$-value distribution is right-skewed. When wrongly specified, this distribution is reversed and becomes left-skew. As such, any data fabrication detected with this method would need to be inspected for misspecified one-tailed hypotheses to preclude false conclusions.

@@

Variance- or standard deviation estimates are typically reported to indicate dispersion, but just like the mean there should be sampling error in this estimate proportional to the sample size [i.e., $\sigma/\sqrt{2n}$ under the assumption of normality, p. 351,@yule1922]. If an observed random variable $x$ is normally distributed, the variance of $x$ is $\chi^2$-distributed [p. 445; @hogg-tanis]; that is
$$
var(x)\sim\frac{\chi^2_{N_j-1}}{N_j-1}
$$
where $N$ is the sample size of the $j$th group. Using the reported summary statistics, we can compute the Mean Squares within ($MS_w$) each condition as
$$
MS_w=\frac{\sum\limits^k_{j=1}(N_j-1)s^2_j}{\sum\limits^k_{j=1}(N_j-1)}
$$
where $s^2_j$ is the variance in the $j$th group. The reported variances can be standardized by dividing the observed variances by the $MS_w$; we denote standardized variances with z^2$.

Standardizing the reported variances allows for several measures of observed dispersion to be calculated. Observed dispersion of the standardized variances can be operationalized as the standard deviation of the variances [denoted in this paper as $SD_z$, @10.1177/0956797613480366] or as the range of the variances (denoted as $max-min_z$). Note that such comparisons are only meaningful if the reported variances originate from a homogeneous population distribution of variances (otherwise, subgroups need to be made).

In order to compute the expected dispersion, we use the distribution of the standardized variances to compute how extreme the observed dispersion of the variances is. Assuming normality, the distribution of the standardized variances follows a $\chi^2$-distribution in the form of
$$
z^2_j\sim\left(\frac{\chi^2_{N_j-1}}{N_j-1}\right)/MS_w
$$
which is the result of the distribution of $var(x)$ as denoted earlier, divided by the $MS_w$. From this distribution, fictitious variances can be generated for each group under investigation that are subsequently used to compute one of the measures for dispersion proposed in the previous paragraph (e.g., standard deviation of the variances). Repeating this across $i$ iterations provides an estimate of the density function for the expection dispersion of the variances, to which the observed dispersion of the variances can be compared. For our purposes, too little dispersion in the variances indicates potential fabrication in the reported data [@10.1177/0956797613480366].

```{r, echo = FALSE}
sd1 <- c(ex1_1esd, ex1_1csd)
sd2 <- c(x2_1esd, x2_2csd)

iterate <- 10000
res1 <- std_var(n = rep(100, length(sd1)), sds = sd1, iter = iterate)
res2 <- std_var(n = rep(100, length(sd2)), sds = sd2, iter = iterate)
```
